---
title: "Colorado Shelter Trends: Multi-Year Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

-   Purpose of this analysis\
-   Data sources and time period covered\
-   Key questions:
    -   Are transfers affecting negative outcomes?
    -   Are adoptions and RTOs keeping pace with intake?
    -   Where are the biggest regional differences?

# Data Overview

## Data Description

-   Overview of data fields (intake, outcomes, transfers, etc.)
-   Data cleaning and preprocessing steps

```{r}
# Load libraries and data
library(tidyverse)
library(readxl)
library(feather)
library(here)

df0 <- read_feather(here("data", "data_transformed", "nkc.feather"))
```

## Summary Statistics

-   Total intakes, adoptions, and outcomes by year

```{r}
df0 %>%
  group_by(year) %>%
  summarise(
    Total_Intake = sum(total_intake, na.rm = TRUE),
    Total_Adoptions = sum(total_adoption, na.rm = TRUE),
    Total_Deaths = sum(total_death, na.rm = TRUE)
  ) %>%
  gt() %>%
  tab_header(
    title = "Shelter Outcomes by Year",
    subtitle = "Total Intake, Adoptions, and Deaths"
  ) %>%
  fmt_number(
    columns = vars(Total_Intake, Total_Adoptions, Total_Deaths),
    decimals = 0
  ) %>%
  cols_label(
    Total_Intake = "Total Intake",
    Total_Adoptions = "Total Adoptions",
    Total_Deaths = "Total Deaths"
  ) %>%
  tab_options(
    table.font.names = "Arial",
    heading.align = "center",
    table.border.top.width = 2,
    table.border.bottom.width = 2
  )
```
