---
title: "Colorado Shelter Trends: Multi-Year Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

-   Purpose of this analysis\
-   Data sources and time period covered\
-   Key questions:
    -   Are transfers affecting negative outcomes?
    -   Are adoptions and RTOs keeping pace with intake?
    -   Where are the biggest regional differences?

# Data Overview

## Data Description

-   Overview of data fields (intake, outcomes, transfers, etc.)
-   Data cleaning and preprocessing steps

```{r}
# Load libraries and data
library(tidyverse)
library(readxl)
shelter_data <- read_excel("/Users/quunc/Downloads/20251001NKCDataSheet.xlsx")
glimpse(shelter_data)
```

## Summary Statistics

-   Total intakes, adoptions, and outcomes by year

```{r}
shelter_data %>%
  group_by(Year) %>%
  summarise(Total_Intake = sum(Intake),
            Total_Adoptions = sum(Adoptions),
            Total_Deaths = sum(Deaths))
```
